Containers and R---the Rockerverse and beyond

### 2020-01-10 manuscript.tex
update manuscript

% !TeX root = RJwrapper.tex
\title{Containers and R: the Rockerverse and beyond}
\author{by Daniel Nüst, Carl Boettiger, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ben Marwick, Karthik Ram, Noam Ross, Nan Xiao, Lori Shepherd, Nitesh Turaga}

@@

The R community keeps growing: the number of new packages on CRAN keeps
on rising, meetups, conferences, online courses and unconferences
prosper, and the uptake in education and industry increases {[}REFs{]}.
All this cements the role of R as the lingua franca of statistics, data
visualisation, and computational research. Coinciding with the rise of R
was the advent of Docker {[}CITATION?{]} as a general tool for
development, distribution, and deployment of software. Combining both
these topics, the \emph{Rocker
Project}(\url{https://www.rocker-project.org/}) was introduced in 2014
and provides a number of
\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} images
for various use cases as described in \citet{RJ-2017-065}. The
considerable uptake and continued evolution of the \emph{Rocker} suite
of containers has lead to numerous projects extending or building upon
\emph{Rocker} images ranging from reproducible research to production
deployments. This article presents this \emph{Rockerverse} of projects.
It also introduces related activities connecting the R language and
environment with other containerisation solutions. The main
contributions is a coherent picture of the current lay of the land of
containers in, with, and for \texttt{R}. This diversity includes
demonstrators, early prototypes, and mature projects.

@@

\emph{Do we need a generic intro here?} DE: Yes, and happy to take a
stab.

\hypertarget{container-images}{%
\section{Container Images}\label{container-images}}

@@

\emph{Microsoft R Open}
(\href{https://github.com/nuest/mro-docker}{MRO}) is an R distribution
formerly known as Revolution R Open (RRO) before Revolution Analytics
was acquired by Microsoft. MRO is compatible with main R and it's
packages.sss ``It includes additional capabilities for improved
performance, reproducibility, and platform support.''
\citep{microsoft_mro} Most notably these capabilities are the MRAN
repository a.k.a. CRAN Time Machine, which is enabled by default, and
the (optional) integration with
\href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel Library}
(MKL) for multi-threading in linear algebra operations
\citep{microsoft_multithread}. MRO does not provide official Docker
images, but a set of community-maintained \texttt{Dockerfile}s and
Docker images \texttt{nuest/mro} are available on Docker Hub at
\url{https://github.com/nuest/mro-docker} and on GitHub at
\url{https://github.com/nuest/mro-docker} respectively. The images are
inspired by the Rocker images and can be used much in the same fashion,
effectively a drop-in replacement allowing users to quickly evaluate if
the benefits of MRO + Intel® MKL apply to their use case. Version-tagged
images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

\emph{Renjin} is an interpreter for the R language running on top of the
\href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java Virtual
Machine} (JVM) providing full two-way access between Java and R code
\citep{wikipedia_renjin_2018}. It was developed to combine the benefits
of R, such scripting and extension packages, with the JVM's advantages
in the areas of security, cross-platform availability, and established
position in enterprise settings. R extension packages need to be
specially compiled and are distributed via the Java package manager
\href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
cf.~\url{http://packages.renjin.org/packages} for available packages.
Packages are loaded on demand, i.e.~at the first call to
\texttt{library()}. Not all R packages, especially one linking to binary
libraries, are available, e.g.~\texttt{rgdal}
\footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.

There are no offical Docker images for Renjin, but community-maintained
images for selected releases only are available under
\texttt{nuest/renjin} on Docker Hub and GitHub at
\url{https://hub.docker.com/r/nuest/renjin} and
\url{https://github.com/nuest/renjin-docker} respectively. These images
expose the command line interface of Renjin in a similar fashion as
Rocker images and allow an easy evaluation of Renjin's suitability, but
are not intended for production use.

\href{http://www.pqr-project.org/}{\emph{pqR}} is a ``a pretty quick
version of R''. \emph{pqR} fixes some opinionated issues in the R
language and is the basis for experimental features, e.g.~automatic
diffentiation\footnote{ \href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development
\href{https://github.com/radfordneal/pqR/}{on GitHub} is a one man
project and it does not provide any Docker images. But especially
disruptive approaches may contribute to the development of the R
ecosystem, so the \texttt{nuest/pqr} images on Docker Hub and GitHub at
\url{https://hub.docker.com/r/nuest/pqr/} and
\url{https://github.com/nuest/pqr-docker} respectively.

\emph{FastR} is ``A high-performance implementation of the R programming
language, built on GraalVM'' (\url{https://github.com/oracle/fastr}). It
is developed by Oracle, connects R to the GraalVM ecosystem
\citep{wikipedia_graalvm_2019}, and also claims superior performance but
also targets full compatibility with base R
\footnote{ \href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.

There are no offical Docker images provided, but
\texttt{nuest/fastr-docker} provides image and \texttt{Dockerfile} on
Docker Hub and GitHub at \href{}{} and
\url{https://github.com/nuest/fastr-docker} respectively.

While the images presented in this section are far from being as vetted,
stable, and widely used as any of the \emph{Rocker} images, they
demonstrate an important advantage of containerisation technology,
namely the abiliy to transparently build portable stacks of open source
software and make them easily accessible to users. All different
distributions are published under GPL licenses. Since all of the
different R distributions claim better performance as a core motivation,
a comparision based on Docker images, potentially leveraging the
\href{https://docs.docker.com/config/containers/resource_constraints/}{resource
restriction mechanisms} of Docker to level the playing field, seems
useful future work.

@@

\href{http://bioconductor.org/help/docker/}{\emph{Bioconductor}} is an
open source, open development project for the analysis and comprehension
of genomic data \citep{gentleman_bioconductor_2004}. The project
consists of 1741 R software packages as of August 15th 2019, as well as
packages containing annotation or experiment data. \_Bioconductor\_has a
semiannual release cycle, each release is associated with a particular
version of R. Docker images allow availability of current and past
versions of \emph{Bioconductor} for convenience and reproducibility.

\emph{Bioconductor} `base' docker images are built on top of
\texttt{rocker/r-ver} and \texttt{rocker/rstudio}. \emph{Bioconductor}
installs packages based on the R version, and therefore uses
\texttt{rocker/rstudio} and \texttt{rocker/r-ver} version tagging.
\emph{Bioconductor} selects the desired version of R from \emph{Rocker},
adds the BiocManager CRAN package for installing appropriate versions of
\emph{Bioconductor} packages, and creates a \emph{Bioconductor} docker
image with an informative tag (R\_version\_Bioc\_version). The images
are summarized on the \emph{Bioconductor} web site
(\url{https://bioconductor.org/help/docker/}), maintained on GitHub
(\url{https://github.com/Bioconductor/bioc_docker}), and available to
the community through \href{https://hub.docker.com/}{DockerHub}. Past
and current combinations of R and \emph{Bioconductor} are therefore
accessible via a specific docker tag.

\emph{Bioconductor} has several images in addition to `base', specific
to various areas of research. The `core' image installs the most
commonly used \emph{Bioconductor} packages. \emph{Bioconductor} images
for proteomics, metabolomics, and flow cytometry are community
maintained. All community maintained images build on top of the
\emph{Bioconductor} base image and therefore indirectly the
\emph{Rocker} images. To simplify building and maintaining
\emph{Bioconductor} images, we use a Ruby templating engine.

A recent audit of our Dockerfiles, following best practices from the
Docker website, led to a reduction in the size and number of layers in
images we produce. The most important insights involve the
\href{https://docs.docker.com/storage/storagedriver/overlayfs-driver/\#how-container-reads-and-writes-work-with-overlay-or-overlay2}{`union'
file system used by Docker}. In this file system, once a layer (e.g.,
\texttt{RUN} statement) writes to a file path, the file path is never
altered. A subsequent layer that might appear to remove or overwrite the
path actually masks, rather than alters, the original. It is therefore
important to clean up (e.g., cache removal) within each layer, and to
avoid re-installing existing dependencies.

@@

\hypertarget{images-for-historic-r-versions-nuestcolinfay-nuest}{%
\subsection{\texorpdfstring{Images for (historic) R versions
{[}@ColinFay,
@nuest{]}}{Images for (historic) R versions , @nuest{[}@ColinFay, @nuest{]}}}\label{images-for-historic-r-versions-nuestcolinfay-nuest}}

\begin{itemize}
\tightlist
\item
  run code with Versions of R: \url{https://srv.colinfay.me/r-online}
\item
  old R versions:
  \url{https://github.com/rocker-org/rocker-versioned/issues/138}
\item
  \href{https://github.com/rocker-org/rocker-versioned/\#version-tags}{semantic
  version tags for Rocker}
\item
  debugging memory problems across multiple versions:
  \url{https://github.com/wch/r-debug}
\item
  cross-R testing: Show how you can run the same script in multiple R
  variants in containers (new contribution). Ideally the images should
  be created using \texttt{containerit} and controlled with
  \texttt{stevedore}/\texttt{docker}, i.e.~via an R script.
\end{itemize}

\hypertarget{windows-images-nuest}{%
\subsection{Windows Images {[}@nuest{]}}\label{windows-images-nuest}}

\begin{itemize}
\tightlist
\item
  \href{https://github.com/nuest/rocker-win}{rocker-win}
\item
  is possible, only relevant in organisations with an existing Windows
  Server-based infrastructure, can meet policies then
\end{itemize}

\hypertarget{non-debian-linux-images-nn}{%
\subsection{Non-Debian Linux images
{[}NN{]}}\label{non-debian-linux-images-nn}}

\begin{itemize}
\tightlist
\item
  Alpine images
\item
  \href{https://github.com/r-hub/rhub-linux-builders}{Images used by
  R-Hub} (overlap with CI?)
\item
  \url{https://github.com/jlisic/R-docker-centos}
\end{itemize}

\hypertarget{r-packages}{%
\section{R Packages}\label{r-packages}}

\hypertarget{packages-for-automation-and-packaging-nuest}{%
\subsection{Packages for automation and packaging
{[}@nuest{]}}\label{packages-for-automation-and-packaging-nuest}}

\begin{itemize}
\tightlist
\item
  \href{https://github.com/o2r-project/containerit/}{\texttt{containerit}}
\item
  \href{https://github.com/traitecoevo/dockertest/}{\texttt{dockertest}}
\end{itemize}

\hypertarget{liftr}{%
\subsubsection{liftr}\label{liftr}}

\href{https://nanx.me/liftr/}{\emph{liftr}} \citep{liftr2019} aims to
solve the problem of persistent reproducible reporting in statistical
computing. Currently, the R Markdown format and its backend compilation
engine \emph{knitr} offer a \emph{de facto} standard for creating
dynamic documents \citep{xie2018}. However, the reproducibility of such
content authoring environments is often limited to individual machines
--- it is not easy to replicate the system environment (libraries, R
versions, R packages) where the document was compiled. This issue
becomes even more serious when it comes to collaborative document
authoring and creating large-scale document building services.
\emph{liftr} solves this reproducibility problem by bringing Docker to
the game. In essence, \emph{liftr} helps R Markdown users create and
manage Docker containers for rendering the documents, thus make the
computations utterly reproducible across machines and systems.

>>>>

% !TeX root = RJwrapper.tex
\title{Containers and R---the Rockerverse and beyond}
\author{by Daniel Nüst, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Karthik Ram, Noam Ross, Nan Xiao, Lori Shepherd, Nitesh Turaga, Jacqueline Nolis, Heather Nolis, Hong Ooi, Ellis Hughes, Ben Marwick}

@@

The R community keeps growing. This can be seen in the number of new
packages on CRAN, which keeps on growing exponentially
\citep{cran:2019}, but also in the numbers of conferences, open
educational resources, meetups, unconferences, and companies taking up,
as exemplified by the useR! conference
series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}},
the global growth of the R and R-Ladies user
groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}},
or the foundation and impact of the
R~Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the \emph{lingua~franca} of
statistics, data visualisation, and computational research. Coinciding
with the rise of R was the advent of
\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
general tool for distribution and deployment of server applications.
Combining both these topics, the \emph{Rocker~Project}
(\url{https://www.rocker-project.org/}) provides images with R (see the
next Section for more details). The considerable uptake and continued
evolution of the Rocker Project has lead to numerous projects extending
or building upon Rocker images, ranging from reproducible research to
production deployments. This article presents this \emph{Rockerverse} of
projects across all development stages: early demonstrations, working
prototypes, and mature products. We also introduce related activities
connecting the R language and environment with other containerisation
solutions. The main contributions is a coherent picture of the current
lay of the land of using containers in, with, and for R.

The article continues with a brief introduction of containerization
basics and the Rocker Project, followed by descriptions of projects
developing images with R installations. Then, we present applications,
starting with the R packages specificly for interacting with Docker,
second-level packages using containers mediately or only for specific
features, up to complex use cases leveraging containers. We conclude
with a reflection on the landscape of packages and applications and
point out future directions of development.

\begin{figure}
\centering
\includegraphics{/home/daniel/git/o2r/rockerverse-paper/rockerverse.png}
\caption{Rockerverse hex sticker}
\end{figure}

@@

Docker, an application and service provide by the eponymous company, has
in just a few short years risen to prominence for development, testing,
deployment and distribution of computer software
\citep[cf.][]{datadog_8_2018,munoz_history_2019}. While there are
related approaches such as
LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}}
or Singularity \citep{kurtzer_singularity_2017}, Docker has become
synomymous with ``containerization''---the method of taking software
artefacts and bundling them in such a way that use becomes standardized
and portable across operationg systems. In doing so, Docker had
recognised and validated the importance of one very important thread
that had been emerging, namely virtualization. By allowing (one or
possibly) multiple applications or services to run concurrently on one
host machine without any fear of interference between them, an important
scalability opportunity is being provided. But Docker improved on
virtualization by accessing the host system---generally Linux---through
a much thinner and smaller shim than a full operating system emulation
or virtualization. This makes for more efficient use of system resources
\citep{felter_updated_2015} and allowed another order of magnitude in
terms of scalability of deployment \citep[cf.][]{datadog_8_2018}. While
Docker makes use of Linux kernel features, its importance was large
enough so that some required aspects of running Docker have been added
to other operating systems to support Docker there more efficiently too
\citep{microsoft_linux_2019}. The success even lead to standardisation
and industry collaboration \citep{oci_open_2019}.

The key accomplishment of Docker (as an ``application'') is to make a
``bundled'' aggregation of software (the so-called ``image'') available
to any system equipped to run Docker, without requiring much else from
the host besides the actual Docker application installation. This is a
rather attractive, and novel, proposition which has lead to widespread
adoption and use of Docker in a variety of domains, \emph{e.g.}, cloud
computing infrastructure \citep{Bernstein2014}, data science
\citep{boettiger_introduction_2015}, and edge computing
\citep{alam_orchestration_2018} {[}more examples!{]}. It provided to be
a natural match for ``cloud deployment'' which runs (or at least appears
to run) ``seamlessly'' without much explicit reference to the underlying
machine, architecture or operating system: containers are portable and
can be deployed with very little in terms of dependencies on the host
system---only the container runtime is required. Images are normally
build from plain text documents called \texttt{Dockerfile}. A
\texttt{Dockerfile} has a specific set of instructions to create and
document a well-defined environment, i.e., install specific software and
expose specific ports.

For statistical computing and analysis centered around R, the Rocker
Project has provided a variety of Docker containers since its start in
2014 \citep{RJ-2017-065}. The Rocker Project provides several lines of
containers spanning to from building blocks with \texttt{R-release} or
\texttt{R-devel}, via containers with RStudio\textasciitilde{}Server and
Shiny\textasciitilde{}Server, to domain-specific containers such as
\texttt{geospatial}. These containers form \emph{image stacks}, building
on top of each other for better maintainability and Also of note is a
series of ``versioned'' containers which match the R release they
contain with the \emph{then-current} set of packages via the MRAN
Snapshot views of CRAN \citep{microsoft_cran_2019}. The Rocker Project's
impact and importance was acknowledged by the Chan Zuckerberg
Initiative's \emph{Essential Open Source Software for Science}, who
provide findung for the projects's sustainable maintenance, community
growth, and targeting new hardware platforms including GPUs
\citep{chan_zuckerberg_initiative_maintaining_2019}.

Docker is not the only containerisation software. An alternative
stemming from the domain of high-perfomance computing is
\textbf{Singularity} \citep{kurtzer_singularity_2017}. Singularity can
run Docker images, and in the case of Rocker works out of the box if the
main process is R, e.g., in \texttt{rocker/r-base}, but does not succeed
in running images where there is an init script, e.g., in containers
that by default run RStudio. In the latter case, a \texttt{Singularity}
file, a recipe akin to a \texttt{Dockerfile}, needs to be used. To date,
no comparable image stack to Rocker exists on
\href{https://singularity-hub.org/}{Singularity Hub}. A further tool for
running containers is
\href{https://github.com/containers/libpod}{\textbf{podman}}, which also
can build \texttt{Dockerfile}s and run Docker images. Proof of concepts
for using podman to build and run Rocker containers
exist\footnote{See \href{https://github.com/nuest/rodman}{https://github.com/nuest/rodman} and \href{https://github.com/rocker-org/rocker-versioned/issues/187}{https://github.com/rocker-org/rocker-versioned/issues/187}}.
Yet the prevelance of Docker, especially in the broader user community
beyond experts or niche systems, and the vast amount of blog posts and
courses for Docker, currently caps specific development efforts for both
Singularity and podman in the R community. This might quickly change
when usability and spread increase, or security features such as
rootless/unprivileged containers, which both these tools support out of
the box, become more sought after.

\hypertarget{container-images}{%
\section{Container images}\label{container-images}}

@@

\textbf{Microsoft R Open}
(\href{https://github.com/nuest/mro-docker}{MRO}) is an R distribution
formerly known as Revolution R Open (RRO) before Revolution Analytics
was acquired by Microsoft. MRO is compatible with main R and its
packages. ``It includes additional capabilities for improved
performance, reproducibility, and platform support.''
\citep{microsoft_mro} Most notably these capabilities are the MRAN
repository, which is enabled by default, and the (optional) integration
with \href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel
Library} (MKL) for multi-threading in linear algebra operations
\citep{microsoft_multithread}. MRO does not provide official Docker
images, but a set of community-maintained \texttt{Dockerfile}s and
Docker images \texttt{nuest/mro} are available on Docker Hub at
\url{https://github.com/nuest/mro-docker} and on GitHub at
\url{https://github.com/nuest/mro-docker} respectively. The images are
inspired by the Rocker images and can be used much in the same fashion,
effectively a drop-in replacement allowing users to quickly evaluate if
the benefits of MRO + Intel® MKL apply to their use case. Version-tagged
images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

\textbf{Renjin} is an interpreter for the R language running on top of
the \href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java
Virtual Machine} (JVM) providing full two-way access between Java and R
code \citep{wikipedia_renjin_2018}. It was developed to combine the
benefits of R, such scripting and extension packages, with the JVM's
advantages in the areas of security, cross-platform availability, and
established position in enterprise settings. R extension packages need
to be specially compiled and are distributed via the Java package
manager \href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
cf.~\url{http://packages.renjin.org/packages} for available packages.
Packages are loaded on demand, i.e., at the first call to
\texttt{library()}. Not all R packages, especially one linking to binary
libraries, are available, e.g., \texttt{rgdal}
\footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained
images for selected releases only are available under
\texttt{nuest/renjin} on Docker Hub and GitHub at
\url{https://hub.docker.com/r/nuest/renjin} and
\url{https://github.com/nuest/renjin-docker} respectively. These images
expose the command line interface of Renjin in a similar fashion as
Rocker images and allow an easy evaluation of Renjin's suitability, but
are not intended for production use.

\textbf{\href{http://www.pqr-project.org/}{pqR}} is a ``a pretty quick
version of R''. \emph{pqR} attempts to improve R on some opinionated
issues in the R language and is the basis for experimental features,
e.g., automatic
diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development
\href{https://github.com/radfordneal/pqR/}{on GitHub} is a one man
project and it does not provide any Docker images. But especially
disruptive approaches may contribute to the development of the R
ecosystem, so the \texttt{nuest/pqr} images on Docker Hub and GitHub at
\url{https://hub.docker.com/r/nuest/pqr/} and
\url{https://github.com/nuest/pqr-docker} respectively.

\textbf{FastR} is ``A high-performance implementation of the R
programming language, built on GraalVM''
(\url{https://github.com/oracle/fastr}). It is developed by Oracle,
connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and
also claims superior performance but also targets full compatibility
with base R
\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
There are no offical Docker images provided, but \texttt{nuest/fastr}
images and \texttt{Dockerfile} are independently maintained on
\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{Docker
Hub} and \href{https://github.com/nuest/fastr-docker}{GitHub}
respectively.

While the images presented in this section are far from being as vetted,
stable, and widely used as any of the Rocker images, they demonstrate an
important advantage of containerisation technology, namely the abiliy to
transparently build portable stacks of open source software and make
them easily accessible to users. All different distributions are
published under GPL licenses. Since all of the different R distributions
claim better performance as a core motivation, a comparision based on
Docker images, potentially leveraging the
\href{https://docs.docker.com/config/containers/resource_constraints/}{resource
restriction mechanisms} of Docker to level the playing field, seems
useful future work.

@@

\href{http://bioconductor.org/}{\emph{Bioconductor}} is an open source,
open development project for the analysis and comprehension of genomic
data \citep{gentleman_bioconductor_2004}. The project consists of 1741 R
software packages as of August 15th 2019, as well as packages containing
annotation or experiment data. \_Bioconductor\_has a semiannual release
cycle, each release is associated with a particular version of R. Docker
images allow availability of current and past versions of
\emph{Bioconductor} for convenience and reproducibility.
\emph{Bioconductor} `base' docker images are built on top of
\texttt{rocker/r-ver} and \texttt{rocker/rstudio}. \emph{Bioconductor}
installs packages based on the R version, and therefore uses
\texttt{rocker/rstudio} and \texttt{rocker/r-ver} version tagging.
\emph{Bioconductor} selects the desired version of R from Rocker, adds
the BiocManager CRAN package for installing appropriate versions of
\emph{Bioconductor} packages, and creates a \emph{Bioconductor} docker
image with an informative tag (R\_version\_Bioc\_version). The images
are summarized on the \emph{Bioconductor} web site
(\url{https://bioconductor.org/help/docker/}), maintained on GitHub
(\url{https://github.com/Bioconductor/bioc_docker}), and available to
the community through \href{https://hub.docker.com/}{DockerHub}. Past
and current combinations of R and \emph{Bioconductor} are therefore
accessible via a specific docker tag.

\emph{Bioconductor} has several images in addition to `base', specific
to various areas of research. The `core' image installs the most
commonly used \emph{Bioconductor} packages. \emph{Bioconductor} images
for proteomics, metabolomics, and flow cytometry are community
maintained. All community maintained images build on top of the
\emph{Bioconductor} base image and therefore indirectly the Rocker
images. To simplify building and maintaining \emph{Bioconductor} images,
we use a Ruby templating engine. A recent audit of the
\emph{Bioconductor} \texttt{Dockerfile}s, following best practices from
the Docker website, led to a reduction in the size and number of layers.
The most important insights involve the
\href{https://docs.docker.com/storage/storagedriver/overlayfs-driver/\#how-container-reads-and-writes-work-with-overlay-or-overlay2}{`union'
file system used by Docker}. In this file system, once a layer (e.g.,
\texttt{RUN} statement) writes to a file path, the file path is never
altered. A subsequent layer that might appear to remove or overwrite the
path actually masks, rather than alters, the original. It is therefore
important to clean up (e.g., cache removal) within each layer, and to
avoid re-installing existing dependencies.

@@

\hypertarget{images-for-historic-r-versions}{%
\subsection{Images for (historic) R
versions}\label{images-for-historic-r-versions}}

As with any other software, each new version of R comes with its share
of changes. Some are breaking changes, some are not, but the fact is
that running a piece of code in a given version might give different
results from another version can be an issue when it comes to
reproducibility of workflows and stability of applications. For example,
think about R random seed: starting with \texttt{R\ 3.6.0}, running
\texttt{set.seed(2811);\ sample(1:1000,\ 2)} will not give the same
result as if it was run inside an older version of R. That might seem
trivial, but all the code using the random number generator will not be
exactly reproducible after this breaking change. Therefore, controlling
the version of software is most crucial for reproducible research
\citep[e.g.][]{boettiger_introduction_2015}.

Containers are perfectly suited to capture a specific configuration of a
computing environment to prohibit such problems. The Rocker images
provides the \emph{versioned stack} for different the version of R since
the project inception, which use the \emph{then-current} stable Debian
base image \citep[cf.][]{RJ-2017-065}. Based on custom build phase
hooks\footnote{\href{https://docs.docker.com/docker-hub/builds/advanced/}{https://docs.docker.com/docker-hub/builds/advanced/}},
i.e., small shell scripts executed at different phases in the automated
build of Rocker images, there are also semantic version tags for the
most recent freezed (i.e., using MRAN) versions with only the major and
minor
version\footnote{See \href{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}{https://github.com/rocker-org/rocker-versioned/blob/master/VERSIONS.md}, FIXME: \href{https://github.com/rocker-org/rocker-versioned/issues/42}{https://github.com/rocker-org/rocker-versioned/issues/42}.}.
So, at the time of writing this article, \texttt{rocker/r-ver:3} and
\texttt{3.6} are aliases for \texttt{3.6.0}, because \texttt{3.6.1} is
the latest release. With the release of \texttt{3.6.2} and pinning of
the MRAN version in \texttt{3.6.1}, the tags \texttt{3} and \texttt{3.6}
will deliver the same image as \texttt{3.6.1}. These tags allow users to
update the base image to retrieve bugfixes while reducing the risk of
also introducing breaking changes.

\href{https://github.com/ColinFay/ronline}{r-online} is an app for
helping users to detect breaking changes between different R versions,
and for historic exploration of R. With a standalone NodeJS app or
\href{https://srv.colinfay.me/r-online}{online}, the user can compare a
piece of code run in two separate versions of R. Internally, r-online
opens one or two Docker instances with the given version of R based on
Rocker images, executes a given piece of code, and returns the result to
the user.

A container also provides an isolated sandbox environment suitable for
testing and evaluation, without interfering with the ``main'' working
environment. This enables cross version testing and bugfixing. To take
this even further, the Rocker contributors are discussing the provision
of R versions reaching back futher than the project's own inception,
reaching back as far as R \texttt{2.x} and \texttt{1.x}
\footnote{\href{https://github.com/rocker-org/rocker-versioned/issues/138}{https://github.com/rocker-org/rocker-versioned/issues/138}}.
The main challenges are finding a suitable base image with a matching OS
version, or making the adjustments to compile R on more recent OS
releases.

\hypertarget{windows-images}{%
\subsection{Windows Images}\label{windows-images}}

Docker containers on the Windows operating system were originally quite
cumbersome to use. They required an extra tool,
\href{https://docs.docker.com/toolbox/}{Docker Toolbox}, which by now
only exists as a legacy solution for older Windows systems. Docker
toolbox leverages \texttt{docker-machine} (see also
Section~\ref{interfaces} for an R package interfacing with
\texttt{docker-machine}) to handle the process of creating a local
virtual host which could host Docker Engine, while exposing the regular
Docker CLI. The Docker CLI commands are forwarded to the virtual host
transparently for the user. For current Windows Server (2016 and later)
and Windows Desktop
(\href{https://www.docker.com/products/docker-desktop}{Docker Desktop}
requires Windows 10) versions, Docker is supported
natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}}
and different base images are offered by
Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
The Docker CLI can be used in just the same way as on other operating
systems but not every base image is supported on every Windows
host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
On Docker Desktop for Windows, the user can run both Linux-based and
Windows-based containers, but only one of the Docker daemons can be used
at a
time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.

\textbf{\href{https://github.com/nuest/rocker-win}{\texttt{rocker-win}}}
is a proof of concept for running R in Windows-based containers
\citep{nust_rocker-win_2019}. It provides selected R versions using
three different base images using \texttt{microsoft/windowsservercore}
and \texttt{mcr.microsoft.com/windows/servercore}. These base images
match different Windows versions: Windows Server 2019, Windows Server
2016, and Windows Server, version 1803. The images are built using two
CI services,
\href{https://docs.travis-ci.com/user/reference/windows/}{Travis CI} and
\href{https://www.appveyor.com/docs/windows-images-software/}{Appveyor},
which provide the different Windows versions and therefore support
different base images, and published automatically
\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker Hub}. The
images are built from manually maintained seperate \texttt{Dockerfile}s,
are tagged with both Windows-variant and R version, e.g.,
\texttt{nuest/rocker-win:ltsc2019-3.6.2} or
\texttt{nuest/rocker-win:1803-latest}, and run the \texttt{R.exe}
process by default. New R versions are only added on demand. The
explorative project demonstrates the variety and pontentially confusing
complexity of Windows on Docker, e.g., because of the different base
images, but also provides a starting point for any Windows-constrained
user to leverage containerisation, including examples for apps using
\textbackslash{}pkg\{plumber, \pkg{Shiny}, or R packages with system
dependencies such as \pkg{sf}. The latter is quite comfortable actually,
as R ships pre-compiled binaries for Windows.

For Windows, just as originally for Linux, the cloud use cases drive the
development and Docker and Microsoft collaborate
closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
In most cases, individual developers or researchers worrying about
reproducibility, will prefer the more widely and host independent
Linux-based containers. Naturally, a Windows license is required for the
host machine. But the developments in the \texttt{rocker-win} prototype
show that containers can also be used for R workflows depending on
Windows-only tools, for leveraging an existing Windows Server-based
infrastructure where policies can otherwise not be met, or for
streamlining interactions with system operating staff that runs
Windows-based containers themselves.

\hypertarget{non-debian-linux-images}{%
\subsection{Non-Debian Linux images}\label{non-debian-linux-images}}

\label{rhub} The \textbf{R-hub} project provides \_``a collection of
services to help R package development'', with the package builder as
the most prominent one\citep{r-hub_docs_2019}. The builder allows R
package developers to check their R package on different platforms and R
versions using a web form or the package \pkg{rhub}
\citep{csardi_rhub_2019}. The builder uses Docker containers to conduct
these checks, taking advantages of their well-defined environments and
sandboxing. The \texttt{Dockerfile}s and images are published on Docker
Hub\footnote{\href{https://hub.docker.com/u/rhub}{https://hub.docker.com/u/rhub}}
and
GitHub\footnote{\href{https://github.com/r-hub/rhub-linux-builders}{https://github.com/r-hub/rhub-linux-builders}}
respectively. The images comprise release and development builds of R
running in the base images of the
\href{https://www.debian.org/}{Debian},
\href{https://ubuntu.com/}{Ubuntu},
\href{https://getfedora.org/}{Fedora}, and
\href{https://centos.org/}{CentOS} linux distributions (Arch Linux is
under
development\footnote{\href{https://github.com/r-hub/rhub-linux-builders/pull/41}{https://github.com/r-hub/rhub-linux-builders/pull/41}},
\texttt{see}rhub::local\_check\_linux\_images()\texttt{for\ a\ full\ list),\ and\ are\ intended\ for\ debugging\ (see\ Section\&nbsp;\textbackslash{}ref\{development\})\ not\ for\ reproducible\ workflows.\ \_The\ rationale\_\ of\ not\ re-using\ the\ Rocker\ image\ stack\ is\ the\ high\ number\ of\ operatings\ systems\ and\ configurations,\ which\ can\ be\ more\ uniformly\ maintained\ in\ a\ seperate\ suite\ of\ images.\ Also,\ a\ common\ structure\ across\ the\ respective\ image\ stacks\ of\ the\ operating\ systems,\ e.g.,\ for}debian\texttt{,\ adding\ GCC\ in}debian-gcc\texttt{and\ then\ R\textquotesingle{}s\ development\ version\ in}debian-gcc-devel`.
The platform also has a specific user configuration differing from the
setup covering typical applications of Rocker images, and the images
include considerably more software, such as a fairly complete LaTeX
installation, to provide an experience closer to CRAN.
